Work Fragmentation in Developer Interaction Data

### 2016-03-19 paper_2.tex
discussion and RQ5

\paragraph{Correlation analysis of the metrics}
%\RR{perhaps expand a bit}
The metrics that characterize the productivity are related to the amount of activity registered and could be possible to find a correlation between them, specially with the number of edits and selections per minute, and the edit ratio. We did a correlation analysis involving these three metrics (Table \ref{tbl:corr_analysis}, classifying the data between traces with or without interruptions. We find that the Edits/min and Selections/minute metrics have low correlation coefficients, which is the expected result as they explain two different kind of activities. The same happens between the Selections/minute and Edit ratio metrics with the data of both kind of traces, this time being a negative correlation which is also an expected value, as the Edit ratio should decrease when the Selections/minute increases.

The Edit/minute and Edit ratio correlation coefficient is higher in both groups, but not high enough to not consider the latter, for it tells something different from the user activity. It is normal to see these two metrics with a moderate correlation coefficient given that the Edit ratio is the proportion of edition events in a minute.

@@

The goal of the exploratory study then was to better understand whether the type of activities taken after an interruption were indicative of different types of activities for each type of session, and also whether these different kinds of activities supported our earlier hypotheses explaining why the two patterns of interruptions were present. Assuming that different commands represented different types of activity, the assumption was that the frequency of a given activity---as represented by the commands used to carry it---would vary between the different types of sessions.

@@

Category \emph{perspective} deals with events related to perspective change in Eclipse, such as switching from the coding to the debugging perspective, or vice-versa. Eclipse allows the user to program in different environments, like enterprise, mobile and Web. It also supports activities regarding other kind of roles, like testing, design and database management. Each of this environments and roles might require to open a different perspective, so having events regarding a perspective change is an indicative of the different roles that a user has.
As such, we have no expectation as to whether this kind of activity is more indicative of program understanding or program editing. % \RR{or do we?}

@@

Category \emph{refactoring} regroups usage of refactoring tools (rename, move methods, extract methods, etc). This is indicative of edition activities in the current piece of code, but it can also affect outer elements and require to verify the changes, implying low and high level navigation. Since automatic refactorings in the IDE are behavior-preserving, the need to verify the changes is lessened. As such, we consider these events to be more indicative of program editing than program comprehension.

@@

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{New category of events. }
	\label{tbl:new_events_udc}
	\centering
	\begin{tabular}{|p{2.5cm}|p{8 cm}| p{2cm} |}
		\hline 
		category & description & frequency \\  
		\hline 
		\hline 
		high-nav & High-level navigation through classes, windows and artifacts & 9.1 \\
		file & Management of files (saving, closing, opening, etc.) &	8.9 \\
		edit-text & Text edition events & 8.1 \\
		debug & Events executed during debugging sessions & 3.3 \\
		text-nav & Low-level text navigation & 3.1 \\
		search & Text or code search & 0.6 \\
		find-replace & Search and replacement of text or code entities & 0.6 \\
		perspective & Change of perspectives in Eclipse & 0.5  \\
		control & Source control events & 0.4 \\
		refactoring & Refactoring events & 0.3\\
		clean-build &  Events related to the build system (clean, build or run a project) & 0.2\\
		\hline
	\end{tabular}
\end{table}

The first thing we noticed is that while some events are used by all users, some events are used heavily by a minority of users and not by others (this matches the observation of Murphy-Hill et al., and by ourselves, that many development tools are under-used \cite{MPB12}). For example, the event \textit{copyLineDown} is more frequent than the \textit{rename} event, but the former is only used by 2.4\% of the users and the latest by 25.2\% of them. As such, we calculated the frequency of event taking into account how commonly used they were.

We calculated the weighted average of every event based on the frequency of execution during the first phase of recovery time after a positive or negative interruption, and the percentage of users that make use of that event (so that events that are used by very few users do not distort the results). The weighted average was calculated as: $$\bar{x} = \frac{\sum_{i=1}^{n}w_ix_i}{\sum_{i=1}^{n}w_i} $$where $x_i$ is de frequency of execution of the event $i$ and $w_i$ is the percentage of users that uses the event. This balances the average of execution for events used by a small number of users.

@@

Summing up, we can conclude that edition-related activities are indeed more frequent after a positive interruption, while comprehension-related activities are more frequent after a negative interruption.  These categories of events agree with our previous hypothesis that positive interruptions are related to information seeking behavior: programmers would interrupt their IDE activities when encountering a roadblock, find the information elsewhere (from the internet, an expert, etc), and return to the IDE with an increased productivity. This also agrees with the literature \cite{PR11, LVD06}. The categories common after nagative interruptions  also agree with our hypothesis that negative interruptions are related to actual task switches where context needs to be rebuilt afterwards, an activity that involves program comprehension \cite{MMLK14, PR12}. %\RR{refs here}.

After a positive interruption, we can see that the more common activities are text edition, refactoring, file management and, to a lesser extent, navigation around perspectives. On the other hand the navigation around classes, find and replace, searches, debug and clean and build are actions more common after a negative one. The navigation in the code and control version activities are more common after the recovery time, having a lower frequency during a positive o negative interruption.

@@

@@

Furthermore, the results indicate a clear diminish of productivity when facing interruptions, but with a high number of interruptions the change of productivity stops decreasing. This could be reflecting what observational studies tell us about multitasking, which is that with higher number of interruptions (increased multitasking) the productivity stops being affected and might increase. The latter actually occurs in the results but the effect is too subtle. 
We intend to reflect the effects of work fragmentation by classifying the interruptions into negative or positive, and in the results we can see a difference in productivity and the kind of activities performed.

@@

We defined several thresholds empirically: 8 hours of inactivity to split a session in sub-sessions, 30 minutes as the minimum duration for a session, 3 minutes for the minimum duration of a interruption and  $\geq$ 12 minutes for prolonged interruptions. We also tested with other close values and obtained similar results. However, a large variation of these thresholds might significantly alter the results.

@@

We analyzed several thousands software evolution tasks, originating from several dozens of developers. Our global analysis found an inverse relationship between number and duration of interruptions and all three of our productivity indicators. These findings agree with the literature on information workers and software developers.

A subsequent local analysis around interruptions confirmed these results as the activity around interruptions was found to be lower than average. This analysis also found two well-defined patterns around interruptions: interruptions yielding a negative effects (consistent with a real-life interruption involving an expensive context switch), and interruptions with positive effects (consistent with information-seeking behavior). Further studies are necessary to expand on these findings.

With the Usage Data Collector dataset, we replicated our previous study to compare the results obtained with Mylyn. We found the same observations on the relationship of interruptions and productivity; even though the observed effects are less marked, we reach similar conclusions.

>>>>

\paragraph{Correlation analysis of the metrics}
%\RR{perhaps expand a bit}
The metrics that characterize the productivity are related to the amount of activity recorded. Thus there is a degree of correlation between them; we study this correlation to determine whether the metrics measure different aspects or not.  %weand could be possible to find a correlation between them, specially with the number of edits and selections per minute, and the edit ratio.

We did a correlation analysis (using Pearson's correlation coefficient) involving these three metrics (Table \ref{tbl:corr_analysis}, classifying the data between traces with or without interruptions. We find that the Edits/min and Selections/minute metrics have a degree of correlation, but that it is relatively low, this is the expected result as they explain two different kind of activities. We do find that the correlation is higher in traces with interruptions. The same happens between the Selections/minute and Edit ratio metrics with the data of both kind of traces, this time being a negative correlation which is also an expected value, as the Edit ratio should decrease when the Selections/minute increases.

The Edit/minute and Edit ratio correlation coefficient is higher in both groups, but not high enough to discard the latter, for it tells something different from the user activity. It is normal to see these two metrics with a moderate correlation coefficient given that the Edit ratio is the proportion of edition events over all events in a minute.

@@

The goal of the exploratory study was to better understand whether the activity taken after an interruption were indicative of different types of behavior for each type of interruption (positive or negative), and also whether these different kinds of activities supported our earlier hypotheses explaining why the two patterns of interruptions were present. Assuming that different commands represented different types of activity, the assumption was that the frequency of a given activity---as represented by the commands used to carry it---would vary between the different types of sessions.

@@

Category \emph{perspective} deals with events related to perspective change in Eclipse, such as switching from the coding to the debugging perspective, or vice-versa. Eclipse allows the user to program in different environments, such as enterprise, mobile and Web. It also supports activities regarding other kind of roles, such as testing, design and database management. Each of this environments and roles might require to open a different perspective, so having events regarding a perspective change is an indicative of the different roles that a user has.
As such, we have no expectation as to whether this kind of activity is more indicative of program understanding or program editing. % \RR{or do we?}

@@

Category \emph{refactoring} regroups usage of refactoring tools (rename, move methods, extract methods, etc). This is indicative of edition activities in the current piece of code, but it can also affect outer elements and requires to verify the changes, implying low and high level navigation. Since automatic refactorings in the IDE are behavior-preserving, the need to verify the changes is lessened. As such, we consider these events to be more indicative of program editing than program comprehension.

@@

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{New category of events. }
	\label{tbl:new_events_udc}
	\centering
	\begin{tabular}{|p{2.5cm}|p{10 cm}| } %p{2cm} |}
		\hline 
		category & description \\ %& frequency \\  
		\hline 
		\hline 
		high-nav & High-level navigation through classes, windows and artifacts \\ %& 9.1 \\
		file & Management of files (saving, closing, opening, etc.) \\ %&	8.9 \\
		edit-text & Text edition events \\ %& 8.1 \\
		debug & Events executed during debugging sessions \\ %& 3.3 \\
		text-nav & Low-level text navigation \\ %& 3.1 \\
		search & Text or code search \\ %& 0.6 \\
		find-replace & Search and replacement of text or code entities \\ %& 0.6 \\
		perspective & Change of perspectives in Eclipse \\ %& 0.5  \\
		control & Source control events \\ %& 0.4 \\
		refactoring & Refactoring events \\ %& 0.3\\
		clean-build &  Events related to the build system (clean, build or run a project) \\ %& 0.2\\
		\hline
	\end{tabular}
\end{table}

First, we noticed that while some events are used by all users, other events are used heavily by a minority of users and not by others; this matches the observation of Murphy-Hill et al. \cite{MBP12}, and by ourselves \cite{KubelkaBCGRRS15}, that many development tools are under-used. For example, the event \textit{copyLineDown} is more frequent than the \textit{rename} event, but the former is only used by 2.4\% of the users and the latest by 25.2\% of them. As such, we calculated the frequency of event taking into account how commonly used they were.

We calculated the weighted average of every event based on the frequency of execution during the first phase of recovery time after a positive or negative interruption, and the percentage of users that make use of that event (so that events that are used by very few users do not distort the results). The weighted average was calculated as: $$\bar{x} = \frac{\sum_{i=1}^{n}w_ix_i}{\sum_{i=1}^{n}w_i} $$where $x_i$ is the frequency of execution of the event $i$ and $w_i$ is the percentage of users that uses the event. This balances the average of execution for events used by a small number of users.

@@

Summing up, we can conclude that edition-related activities are indeed more frequent after a positive interruption, while comprehension-related activities are more frequent after a negative interruption.  These categories of events agree with our previous hypothesis that positive interruptions are related to information seeking behavior: programmers would interrupt their IDE activities when encountering a roadblock, find the information elsewhere (from the internet, an expert, etc), and return to the IDE with an increased productivity. This also agrees with the literature \cite{PR11, LVD06}. The categories common after negative interruptions  also agree with our hypothesis that negative interruptions are related to actual task switches where context needs to be rebuilt afterwards, an activity that involves program comprehension \cite{MMLK14, PR12}. %\RR{refs here}.

%After a positive interruption, we can see that the more common activities are text edition, refactoring, file management and, to a lesser extent, navigation around perspectives. On the other hand the navigation around classes, find and replace, searches, debug and clean and build are actions more common after a negative one. The navigation in the code and control version activities are more common after the recovery time, having a lower frequency during a positive o negative interruption.

@@

\paragraph{Discussion of the thresholds.} We used several thresholds in this study; we regroup their motivations in the following.

Perhaps the most important thresholds are the inactive time thresholds that delimit interruptions: a \emph{short interruption} is detected after an inactivity of 3 minutes, while a \emph{long interruption} is detected after 12 minutes. Both values come from the literature. More specifically, the work of Gonzalez and Mark \cite{GM04} found that interruptions lasted around 3 minutes, in a field study, while Kaptelinin and Nardi \cite{KaptelininN07} found that interruptions of more than 12 minutes were more likely to corespond to switching activity on a different project or topic.

We classified sessions in groups based on the empirical distribution of the interruptions. In particular, to study the relationship between the number of interruptions and productivity, we split each dataset in four groups based on the quartiles and the median, in order to study the evolution as the number of interruptions grow. We used empirically-defined thresholds since we did not find similar thresholds in the literature. %In the absence of reference from the literature on this topic, basing ourselves on the distribution is the most reasonable choice we could make.

We studied the impact of the duration of interruptions similarly, by computing the ratio of long interruptions over all interruptions, and creating groups based on whether the interruptions in a session were predominantly short or long, for a similar reason.

The duration of the recovery time that we chose, 15 minutes, is the one mentioned by DeMarco and Lister in Peopleware \cite{demarco1999peopleware}. The division in 5 minutes chunks to study the recovery in more details is a compromise between the resolution of the analysis, and the amount of noise. On a related note, we filtered sessions so that they all have at least a duration of 30 minutes, so that sessions have enough data for the analysis to be meaningful; we argue that having less than twice the recovery time is simply not enough for a session to be meaningful. Similarly, in our local analysis, we filter sessions that have less than 30 minutes before or after an interruption so that we can study both the time during and after recovery, without having this period impacted by another interruption.

%\textbf{
%Interruptions: at least 3 minutes

%Prolonged interruptions: at least12 minutes

%Num. of interruptions: 
%    -mylyn: none, [1-3], [4-5], [6-10], [>11] (minutes)
%     -udc: none, [1-2], [3-4], [5-7], [>7] (minutes

%Proportion of prol. interruptions: 0.50

%Recovery time: 15 min, split into 5 minutes phases

%Before interruption: 10 minutes before the next interruption

%Local analysis: 30 minutes around an interruption

%Filtering sessions: at least 30 minutes of duration

%Creo que la idea era argumentar (en la respuesta a las revisiones) que explorar otras posibilidades en estos thresholds implica mucho trabajo y experimentos, corrígeme si estoy equivocado.

@@

Our results indicate a clear drop of productivity when facing interruptions, but with a high number of interruptions the change in productivity stops decreasing. This could be reflecting what observational studies tell us about multitasking, which is that with higher number of interruptions (increased multitasking) the productivity stops being affected and might increase. %The latter actually occurs in the results but the effect is too subtle. 
We inspected  the effects of work fragmentation in more details by classifying the interruptions into negative or positive ones, and in the results we can see a difference in productivity and the kind of activities performed.

@@

We defined several thresholds empirically: 8 hours of inactivity to split a session in sub-sessions, 30 minutes as the minimum duration for a session, 3 minutes for the minimum duration of a interruption and  $\geq$ 12 minutes for prolonged interruptions. We explained our motivations for these and other thresholds in the discussion. We also tested with other close values and obtained similar results. However, a large variation of these thresholds might alter the results.

@@

Based on the Mylyn dataset, we analyzed several thousands software evolution tasks, originating from several dozens of developers. Our global analysis found an inverse relationship between number and duration of interruptions and all three of our productivity indicators. These findings agree with the literature on information workers and software developers.

A subsequent local analysis around interruptions confirmed these results as the activity around interruptions was found to be lower than average. This analysis also found two well-defined patterns around interruptions: interruptions yielding a negative effects (consistent with a real-life interruption involving an expensive context switch), and interruptions with positive effects (consistent with information-seeking behavior). %Further studies are necessary to expand on these findings.

With the Usage Data Collector dataset, we replicated our previous study to compare the results obtained with Mylyn. We found the same observations on the relationship of interruptions and productivity; even though the observed effects are less marked, we reached similar conclusions.
